Qualifying Exam: Examples of my approach and research

### 2017-12-07 doc/qualifying_exam/vpd.tex
think done with qualifyin exam presentation

\title[Your Short Title]{When does vapor pressure deficit drive or reduce evapotranspiration?}
\author{Adam Massmann,  Pierre Gentine and Changjie Lin}
\institute{AGU Fall Meeting (Rough Draft)}
\date{December 14th, 2017}

@@

\section{Introduction}
\begin{frame}{Does VPD drive or reduce ET? - hydrometeorologists' perspective}
  \begin{Huge}
  \[VPD = (1-RH)\cdot e_s (T)\]
\end{Huge}
  \begin{itemize}
  \item \textbf{Hydrometeorologists} would say that an increase in VPD (increase in \textbf{atmospheric demand}) would drive an \textbf{increase in ET}.
  \end{itemize}
\end{frame}

\begin{frame}{Does VPD drive or reduce ET? - physiologists' perspective}
  \begin{figure}
  \includegraphics[width=2.25in]{stomata.png}% was O(1.5in)
  \caption{from evolution.berkeley.edu}
\end{figure}
  \begin{itemize}
  \item However, \textbf{plant physiologists} know that plants have evolved to use stomata to conserve and regulate water use. So \textbf{stomata closure} in response to increases in VPD may \textbf{decrease ET}.
  \end{itemize}
\end{frame}

\begin{frame}{The question is, which effect dominates with an increase in VPD: plant response (decrease in ET) or atmospheric demand (increase in ET)?}
  \begin{itemize}
  \item It should be a function of plant type and the environment:
    \begin{itemize}
    \item Plants that are evolved to conserve water will tend to reduce ET with increases in VPD.
    \item However the environment can overwhelm plant response: at some threshold (i.e. very high VPD) the atmospheric demand should dominate and plants will not be able to conserve water, no matter how much they have evolved to do so.
    \end{itemize}
  \end{itemize}
\end{frame}

\section{Method}
\begin{frame}{Use physically reasonable assumptions to develop a theory.}
  We can use Penman-Monteith (PM) to estimate ET:
  \[ET = \frac{\Delta R + g_a \rho_a c_p \, VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},\]
\textbf{Problem}: $g_s$ (stomatal conductance) is a function of photosynthesis, which is a function of ET itself.  So ET in PM is really an implicit function of itself and we cannot take derivatives!
\end{frame}

\begin{frame}{Use physically reasonable assumptions to develop a theory.}
Apply a constant uWUE assumption (conserved within plant type; see \cite{Zhou_2016}):
\[uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET},\]
To derive a new form of Penman-Monteith without implicit ET dependence:
  \[  ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p \, VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6 \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right)\footnote{note that all terms are known }}{ \Delta + \gamma}\]
\end{frame}

\begin{frame}{Use physically reasonable assumptions to develop a theory.}
  With our new form of Penman-Monteith we can now take derivatives, giving:
  \[\frac{\partial \;  ET}{\partial \, VPD} = \frac{2 g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R*\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)\]
  In the interest of time, we will just focus in the ``sign'' term:
  \[\sign \left[\frac{\partial \;  ET}{\partial \, VPD}\right] = \sign \left[  \left(\frac{ c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R*\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right) \right] \]

\end{frame}

\section{Results - theory}
\begin{frame}{Consequences of theory - ``sign'' term}
  \[\frac{\partial \, ET}{\partial \, VPD} = \text{scaling} \cdot \left(\frac{ c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R*\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)\]
  \begin{itemize}
  \item $c_p$ and $R*$ are constants
  \item $R_{air}$, $\gamma$, and $c_s$ are approximately constant (relative to $\sqrt{VPD}$)
  \item uWUE and $g_1$ are constants within plant type (i.e. grass, crops, deciduous forest, evergreen needleleaf forest, shrub)
  \item Which just leaves VPD
  \end{itemize}
  So within each plant type, whether the atmospheric demand (ET increasing with VPD) or plant response (ET decreasing with VPD) dominates is essentially just a function of VPD!
\end{frame}

\begin{frame}{``Sign'' Term}
  \begin{figure}
    \adjincludegraphics[width=3.5in, trim={0 {0.51\height} 0 0}, clip]{fig05.pdf}
  \end{figure}
\end{frame}

\section{Test Theory}
\begin{frame}{The theory seems nice, but we should test with data!}
    \begin{itemize}
    \item Introduce a free uncertainty parameter $\sigma$ to Penman Monteith:
      \[ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6\; \boldsymbol{\sigma} \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}\]
    \item At each observation from FLUXNET (56 sites across globe) calculate a unique $\sigma$:
      \small
      \[\sigma = - \frac{g_a \gamma c_s \sqrt{VPD} P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})}\]
      \normalsize
    \item Then propagate uncertainty forward by including $\sigma$ in the derivative:
    \end{itemize}
      \[\frac{\partial \;  ET}{\partial \; VPD} = \text{scaling} \cdot \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)\]

  \end{frame}

@@

    \begin{frame}{Test theory with FLUXNET data - ENF}
                \includegraphics[width=3.5in]{enf_eee.png}
     \end{frame}

@@

    \begin{frame}{Test theory with FLUXNET data - GRA}
                \includegraphics[width=3.5in]{gra_eee.png}
     \end{frame}

% \begin{frame}{Summary of theory}
%   \begin{itemize}
%   \item We use \cite{Zhou_2016}'s uWUE to derive a new analytically tractable form of PM.
%   \item This new analysis suggests that the ``tipping'' point for which atmospheric demand overwhelms plant response will be almost exclusively a function of VPD.
%     \begin{itemize}
%     \item For each PFT there will be a VPD$_{crit}$ above which atmospheric demand will dominate and ET will increase with VPD.
%     \end{itemize}
%   \item Plant types evolved to conserve water (CSH) have a higher VPD$_{crit}$ than plants evolved (or bred) to use water and prioritize GPP (CRO). Trees and grasslands are somewhere between these two extremes.
%   \item Aerodynamic conductance scales the response, so plants with large surface roughness will be more likely to have a larger response as there is less resistance between the surface and the atmosphere.
%   \end{itemize}
% \end{frame}

\section{Conclusions}
% \begin{frame}{Summary}
%   \begin{itemize}
%   \item Theory finds that the ``tipping'' point for which atmospheric demand overwhelms plant response will be almost exclusively a function of VPD.  Plant types evolved to conserve water (CSH) have a higher VPD$_{crit}$ (and more negative ET response) than plants evolved (or bred) to use water and prioritize GPP (CRO).
%   \item On average, ecosystem response to VPD follows roughly what we might expect: CRO (prioritize GPP) has positive ET response to VPD, while all others have a negative response. Ordering by increasing magnitude of negative response gives: DBF, GRA, ENF, CSH; which roughly correspond to expectations for increasing water conservation as a function of PFT.
%   \item Uncertainty is high, especially for CRO and GRA. However, inclusion of uncertainty does not change the story for ENF or CSH.
%   \end{itemize}
% \end{frame}

\begin{frame}{Summary}
  \small
  \begin{itemize}
  \item Derived a new form of Penman-Monteith that explicitly calculates ET given published constants.
  \item Developed a theory for when atmospheric dryness will increase ET (atmospheric demand dominates) vs. decrease ET (plant response dominates). Theory tests well using real data from FLUXNET2015 on plant types CSH, ENF, and DBF. Applicability to CRO and GRA is more uncertain.
  \item For plants evolved to prioritize water conservation over production, atmospheric demand is less likely to dominate.
  \item All plant types except for crops had common occurrence of decreases in ET with increases in VPD (plant response dominates). Therefore, the hydrometeorologists perspective is often wrong, and drought indices built upon assumptions about atmospheric demand are unrealistic for vegetated surfaces.
  \item Our form of Penman-Monteith is capable of estimating ET realistically and could be used as a replacement for PET in drought indices (e.g. Palmer Drought Severity Index - PDSI).
  \end{itemize}
\end{frame}

>>>>

\title[Your Short Title]{Qualifying Exam: Examples of my approach and research}
\author{Adam Massmann}
\date{December 7th, 2017}

@@

\section{Introduction}
\begin{frame}{Current project - When does vapor pressure deficit (VPD) drive or reduce evapotranspiration ET?}
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.5\textwidth}
      \begin{minipage}{\textwidth}

        \begin{itemize}
        \item A \textbf{hydrometeorologist} might say that an increase in VPD (increase in \textbf{atmospheric demand}) would drive an \textbf{increase in ET}.
        \item However, \textbf{plant physiologists} know that plants have evolved to use stomata to conserve and regulate water use. So \textbf{stomata closure} in response to increases in VPD may \textbf{decrease ET}.
        \end{itemize}
      \end{minipage}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{minipage}{\textwidth}
        \begin{figure}
          \includegraphics[width=2.00in]{stomata.png}% was O(1.5in)
          \caption{from evolution.berkeley.edu}
        \end{figure}
      \end{minipage}
    \end{column}
  \end{columns}
\end{frame}

\section{Method}

\begin{frame}{My approach}
  \begin{itemize}
  \item Use physically reasonable assumptions guided by intuition and literature to simplify the problem and develop a transparent and intuitive theory (or hypothesis). The goal is to capture the leading order behavior of the system.
    \begin{itemize}
    \item Here we start with Penman Monteith, and apply an assumption of constant uWUE (recently identified by \cite{Zhou_2016}) which gives a differentiable function of ET.
    \end{itemize}
  \item Test theory with observational or sophisticated model data to see if theory is valid, and to what extent it may generalize.
    \begin{itemize}
    \item Here we use FLUXNET2015 data.
    \end{itemize}
  \end{itemize}
\end{frame}

\section{Results - theory}
\begin{frame}{Results - theory's form for ET response to VPD perturbations}
  \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] (image) at (0,0) {\adjincludegraphics[width=3.5in, trim={0 {0.51\height} 0 0}, clip]{fig05.pdf}};
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
        %% next four lines will help you to locate the point needed by forming a grid. comment these four lines in the final picture.↓
       % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
       % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
       % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
       % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
       %% upto here
        \draw[-latex, line width=2pt] (0.83,0.725) --++(0.0in,0.3in)node[anchor=south] {ET Increasing};
        \draw[-latex, line width=2pt] (0.83,0.725) --++(0.0in,-0.3in)node[anchor=north] {ET Decreasing};

        % \draw[dashed,-latex, line width=5pt] (0.8,0.75) --++(0.0in,+0.5in)node[anchor=north] {ET Decreasing};
    \end{scope}
    % \node[align=center,red,font={\large}] at (image.southwest) {ET decreasing};
  \end{tikzpicture}
\end{frame}

\section{Test Theory}
\begin{frame}{Testing Theory - the good: evergreen forest}
  \includegraphics[width=3.5in]{enf_eee.png}
\end{frame}

\begin{frame}{Testing Theory - the bad: grassland}
  \includegraphics[width=3.5in]{gra_eee.png}
\end{frame}

\section{Conclusions}

\begin{frame}{Summary - When does VPD reduce ET?}
  \small
  \begin{itemize}
  \item Successfully simplified the problem to develop a theory for when atmospheric dryness will increase ET (atmospheric demand dominates) vs. decrease ET (plant response dominates).
  \item Tested the theory with real data from FLUXNET2015: shrubs and forests tested well, while the theory's applicability to crops and grass is more uncertain.
  \item All plant types except for crops has a common occurrence of decreases in ET with increases in VPD (plant response dominates). Therefore, any drought indices which assume that atmospheric demand dominates (e.g PET-based metrics) are likely unrealistic for vegetated surfaces.
  \end{itemize}
\end{frame}

@@

@@
